Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-14 paper/main.tex
more rnaseq res

\title[short Title]{Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning}
\author[Sample \textit{et~al}.]{Felix Stiehler\,$^{\text{\sfb 1,}}$$^\dagger$, Alisandra Denton\,$^{\text{\sfb 1,}}$$^\dagger$\, Marvin Steinborn\,$^{\text{\sfb 1}}$, Stephan Scholz, Andreas Weber\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Institue for Plant Biochemistry, Heinrich-Heine-University, Dusseldorf, D-40225, Germany}

@@

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, Augustus predictions and the
reference to RNAseq coverage data. RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences \ref{fig:examples} % and supplemental figures.
we identified cases where RNAseq supported both the annotations of the reference 
and Helixer (a), neither (b), the reference, but not Helixer (c),
and Helixer, but not the reference (d). Helixer models do not yet have postprocessing
to make finalized single predictions, but instead output basewise probabilities.
We see that the model exibits higher uncertainty around transitions between annotation `
classes, for instance between UTR and CDS, or more dramatically between UTR and 
intergenic even where Helixer predictions closely match the reference and RNAseq data (a).

In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (b and c,hSlA2 supp). However, in the extreme,
there are cases where the Helixer model exibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (hSlA2, hSlA3, lSlA3). In one case the 
has not made a reasonably is not assigning a sufficiently 
confident probability to any

\section{Discussion}
We propose a novel way to develop gene calling models that are able to perform 
cross-species genomic annotiation for a wide variety of large eucaryotic genomes. 
Helixer does this while outperforming AUGUSTUS on a base pair wise level and 
withstanding RNASeq data based scrutiny.

>>>>

\title[short Title]{Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning}
\author[Sample \textit{et~al}.]{Felix Stiehler\,$^{\text{\sfb 1,}}$$^\dagger$, Alisandra Denton\,$^{\text{\sfb 1,}}$$^\dagger$\, Marvin Steinborn\,$^{\text{\sfb 1}}$, Daniela Dey, Stephan Scholz, Andreas Weber\,$^{\text{\sfb 1,}*}$}
\address{$^{\text{\sf 1}}$Institue for Plant Biochemistry, Heinrich-Heine-University, Dusseldorf, D-40225, Germany}

@@

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, Augustus predictions and the
reference to RNAseq coverage data. RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences \ref{fig:examples} % and supplemental figures.
we identified cases where RNAseq supported both the annotations of the reference 
and Helixer (a), neither (b), the reference, but not Helixer (c),
and Helixer, but not the reference (d).

Helixer models do not yet have postprocessing
to make finalized single predictions, but instead output basewise probabilities.
We see that the model exibits higher uncertainty around transitions between annotation 
classes, for instance between UTR and CDS, or more dramatically between UTR and 
intergenic even where Helixer predictions closely match the reference and RNAseq data (a). 
Helixer's uncertainty around UTR <-> intergenic transitions may relate to a
fundamentally harder problem (there is no conserved motif at the site as is observed
for splice sites and start/stop codons), lack a single precise biological site \citep{},
or noise in the reference, which we observed relative to the RNAseq data \ref{}. % supplemental UTR figures, 1/species

Helixer models also sometimes showed uncertainty for larger regions.
In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (b and c,hSlA2 supp). However, in the extreme,
there are cases where the Helixer model exibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (hSlA2, hSlA3, lSlA3). Notably, in one of the examples ()
the RNAseq shows evidence of alternative splicing; and in another () Helixer's
prediction falls between that of the reference and Augustus.

Coverage and Spliced coverage were broken down by the confusion matrix (of both
Helixer vs Augustus and Helixer vs the reference) for all genomes \ref{}. % H vs A main, H vs R supp
Where tools were in agreement (far left), coverage and spliced coverage closely matched
expectations. Specifically for CDS:CDS and UTR:UTR most base pairs had some, and many
had moderate or high coverage. The same pattern was seen for intron:intron and spliced
coverage. Finally, intergenic:intergenic showed only a small fraction of base pairs with 
any of either coverage or spliced coverage.
In cases of conflict, Helixer's predictions were substantially more consistent with
the RNAseq data than those of Augustus. Specifically, for base pairs where one tool said CDS and
the other intergenic or intron \ref{}, there was more coverage when 
Helixer said CDS. Similarly, for base pairs where one tool said intron and the other 
CDS or intergenic, there was more spliced coverage when Helixer said intron \ref{}. 
These patterns were consistent in direction but varied in magnitude in the individual species. % double check
RNAseq based comparison of Helixer and the reference was less clear cut.
-Helixer is good at differentiating ig, ref better at introns
-diffs between species. 
-Helixer is substantially better for Tcacao, with a "poor Sub Genic F1"
-taken together: competitive quality predictions, in some cases the mistakes really are in the reference / we wouldn't want 100\% accuracy against the reference

\section{Discussion}
We propose a novel way to develop gene calling models that are able to perform 
cross-species genomic annotiation for a wide variety of large eucaryotic genomes. 
Helixer does this while outperforming AUGUSTUS on a base pair wise level and 
consistency with independent RNASeq data.
